---
title: "CodeBook"
author: "David Kessler"
date: "10/21/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## CodeBook for David Kessler's Coursera Getting and Cleaning Data course project

This is the CodeBook for the Coursera Getting and Cleaning Data course project. Analysis steps in this document will have numbers following by a period. As in "1." below. All of text here is also contained as comments in the run_analysis R script, which is included as the singular code chunk below. The steps seem to me to flow one to another to another, so I didn't do intermediate result displays, etc.

1. Check if the tidyverse package is installed and install if it isn't
invisible(if(!is.element("tidyverse", installed.packages()[,1])){install.packages("tidyverse")})
2. load the tidyverse library  
library(tidyverse)
3. Establish the location of the initial data zip folder 
zipurl <- "https://d396qusza40orc.cloudfront.net/getdata%2Fprojectfiles%2FUCI%20HAR%20Dataset.zip"
4. Store the current working directory as a text string for use in navigating to various folders 
startingwd <- getwd()
5, Store the future zip file name 
zipfile <- "getdata_projectfiles_UCI HAR Dataset.zip"
6. Store the location of the zip file 
zipcombine <- as.character(paste(startingwd, zipfile, sep = "/"))
7. If the file exists, remove it 
invisible(if(file.exists(zipcombine)) {file.remove(zipcombine)})
8. Store the directory location for the unzipped folder
datadir <- paste0(startingwd, "/UCI HAR Dataset")
9. If the folder exists, remove it
invisible(if(dir.exists(datadir)) {unlink(datadir, recursive = TRUE)})
10. Download and store the zip file 
download.file(zipurl, destfile = zipcombine)
(It appears one does not have to be logged in to coursera.org to complete this step)
11. Unzip the file 
unzip(zipfile)
12. store the measurment names 
measurementnames <- read_table(paste0(datadir, "/features.txt"), col_names = FALSE)
13. transpose the measurement names so that it is 1x561 instead of 561x1 
measurementnames <- t(measurementnames)
14. read in the test set measurements  
xtest <- read_delim(paste0(datadir, "/test/X_test.txt"), delim = " ", col_names = FALSE)
15. update the measurements in the test set from character to numeric 
xtest <- xtest %>% mutate_if(is.character,as.numeric)
16. name the xtest columns 
colnames(xtest) <- measurementnames
17. read in the test set activity codes 
testactivities <- read_table(paste0(datadir, "/test/y_test.txt"), col_names = FALSE)
18.read in the test set subjects 
testsubjects <- read_table(paste0(datadir, "/test/subject_test.txt"), col_names = FALSE)
19. Read in the activity labels 
activitylabels <- read_table(paste0(datadir, "/activity_labels.txt"), col_names = FALSE)
20. Name the activity labels column as activity 
colnames(activitylabels) <- c("activityindicator", "activityname")
21. Name the test set activities column 
colnames(testactivities) <- c("activityindicator")
22. merge the activity labels into the test set activities 
testactivities <- merge(testactivities, activitylabels)
23. Column bind the test set subjects and activities to the test set measurements 
xtest <- cbind(testsubjects, testactivities, xtest)
(Now we're going to do the similar steps for the training data set.)
24. read in the training set measurements 
xtrain <- read_delim(paste0(datadir, "/train/X_train.txt"), delim = " ", col_names = FALSE)
25. update the measurements in the training set from character to numeric 
xtrain <- xtrain %>% mutate_if(is.character,as.numeric)
26. name the xtrain columns 
colnames(xtrain) <- measurementnames
27. read in the training set activity codes 
trainactivities <- read_table(paste0(datadir, "/train/y_train.txt"), col_names = FALSE)
28. read in the test set subjects 
trainsubjects <- read_table(paste0(datadir, "/train/subject_train.txt"), col_names = FALSE)
29. Name the training set activities column 
colnames(trainactivities) <- c("activityindicator")
30. merge the activity labels into the training set activities 
trainactivities <- merge(trainactivities, activitylabels)
31. Column bind the training set subjects and activities to the training set measurements 
xtrain <- cbind(trainsubjects, trainactivities, xtrain)
32. Combine the test and train data sets into one table 
alldata <- rbind(xtrain, xtest)
33. Determine the locations of the mean and standard deviation variables 
meanandstdlocations <- intersect(grep("mean|std", measurementnames), grep("meanFreq", measurementnames, invert = TRUE))
(The above intentionally excludes the meanFreq measurements, as those are "Weighted average of the frequency components to obtain a mean frequency", which is distinct from the mean, which is Mean value, and is a member of the set for the same measurements such as Body Acceleration Jerk and Body Gyro.)
34. Create a reduced data out the subject and activity columns in preparation to reduce to only the mean and standard deviation measurements 
meanandstddata <- alldata[,-c(1:3)]
35. Create a reduced data set containing only the mean and standard deviation measurements 
meanandstddata <- meanandstddata[,meanandstdlocations]
36. Column bind the subjects and activities to the mean and stadard deviation measurements 
meanandstddata <- cbind(alldata[,c(1,3)], meanandstddata)
37. Rename the first column in the mean and stadard deviation data set to a descriptive name 
colnames(meanandstddata)[1] <- "subject"
38. Store the mean and stadard deviation data set column names to a vector 
thenames <- colnames(meanandstddata)
39. Remove the leading digits and spaces from the names vector 
thenames <- gsub("\\d+ ","", thenames)
40. Remove the opening parentheses from the names vector 
thenames <- gsub("\\(+","", thenames)
41. Remove the closing parentheses from the names vector 
thenames <- gsub("\\)+","", thenames)
42. Remove the dashes from in the from the names vector 
thenames <- gsub("-","", thenames)
43. Substitute "acceleration" for "Acc: in the names vector 
thenames <- gsub("Acc","acceleration", thenames)
44. Change the names vector to all lowercase (a preference listed durign the lecture by Dr. Leek) 
tolower(thenames)
45. Rename the columns in the mean and stadard deviation data set 
colnames(meanandstddata) <- thenames
46. create a tidy data set containing the average of each variable for each activity and each subject 
averagesdata <- meanandstddata %>% group_by(activityname, subject) %>% summarise_all("mean")
47. Extract the variable names in the averages data set 
averagesnames <- colnames(averagesdata)
48. Append "average" to each of the averages data names items 
averagesnames <- paste0(averagesnames,"average")
49. Remove the "average" from activity name and subject, as these are not averages 
averagesnames[1:2] <- c("activityname", "subject")
50. Update the variable names in the averages data set 
colnames(averagesdata) <- averagesnames
51. check if the final step output file already exists, if so delete it
finalfile <- as.character(paste(startingwd, "measurementaveragesbysubjectandactivity.csv", sep = "/"))
invisible(if(file.exists(finalfile)) {file.remove(finalfile)})
52. Write the final tidy data set to a csv and also View
write.csv(averagesdata, file = "measurementaveragesbysubjectandactivity.csv")
View(averagesdata)

```{r r script run_analysis.R, echo= FALSE}
#check if the tidyverse package is installed and install if it isn't
invisible(if(!is.element("tidyverse", installed.packages()[,1])){install.packages("tidyverse")})

# load the tidyverse library  
library(tidyverse)

# Establish the location of the initial data zip folder 
zipurl <- "https://d396qusza40orc.cloudfront.net/getdata%2Fprojectfiles%2FUCI%20HAR%20Dataset.zip"

# Store the current working directory as a text string for use in navigating to various folders 
startingwd <- getwd()

# Store the future zip file name 
zipfile <- "getdata_projectfiles_UCI HAR Dataset.zip"

# Store the location of the zip file 
zipcombine <- as.character(paste(startingwd, zipfile, sep = "/"))

# If the file exists, remove it 
invisible(if(file.exists(zipcombine)) {file.remove(zipcombine)})

# Store the directory location for the unzipped folder
datadir <- paste0(startingwd, "/UCI HAR Dataset")

# If the folder exists, remove it
invisible(if(dir.exists(datadir)) {unlink(datadir, recursive = TRUE)})

# Download and store the zip file 
download.file(zipurl, destfile = zipcombine)
# (It appears one does not have to be logged in to coursera.org to complete this step)

# Unzip the file 
unzip(zipfile)

# store the measurment names 
measurementnames <- read_table(paste0(datadir, "/features.txt"), col_names = FALSE)

# transpose the measurement names so that it is 1x561 instead of 561x1 
measurementnames <- t(measurementnames)

# read in the test set measurements  
xtest <- read_delim(paste0(datadir, "/test/X_test.txt"), delim = " ", col_names = FALSE)

# update the measurements in the test set from character to numeric 
xtest <- xtest %>% mutate_if(is.character,as.numeric)

# name the xtest columns 
colnames(xtest) <- measurementnames

# read in the test set activity codes 
testactivities <- read_table(paste0(datadir, "/test/y_test.txt"), col_names = FALSE)

# read in the test set subjects 
testsubjects <- read_table(paste0(datadir, "/test/subject_test.txt"), col_names = FALSE)

# Read in the activity labels 
activitylabels <- read_table(paste0(datadir, "/activity_labels.txt"), col_names = FALSE)

# Name the activity labels column as activity 
colnames(activitylabels) <- c("activityindicator", "activityname")

# Name the test set activities column 
colnames(testactivities) <- c("activityindicator")

#21. merge the activity labels into the test set activities 
testactivities <- merge(testactivities, activitylabels)

#22. Column bind the test set subjects and activities to the test set measurements 
xtest <- cbind(testsubjects, testactivities, xtest)

## Now we're going to do the similar steps for the training data set.

# read in the training set measurements 
xtrain <- read_delim(paste0(datadir, "/train/X_train.txt"), delim = " ", col_names = FALSE)

# update the measurements in the training set from character to numeric 
xtrain <- xtrain %>% mutate_if(is.character,as.numeric)

# name the xtrain columns 
colnames(xtrain) <- measurementnames

# read in the training set activity codes 
trainactivities <- read_table(paste0(datadir, "/train/y_train.txt"), col_names = FALSE)

# read in the test set subjects 
trainsubjects <- read_table(paste0(datadir, "/train/subject_train.txt"), col_names = FALSE)

# Name the training set activities column 
colnames(trainactivities) <- c("activityindicator")

# merge the activity labels into the training set activities 
trainactivities <- merge(trainactivities, activitylabels)

# Column bind the training set subjects and activities to the training set measurements 
xtrain <- cbind(trainsubjects, trainactivities, xtrain)

# Combine the test and train data sets into one table 
alldata <- rbind(xtrain, xtest)

# Determine the locations of the mean and standard deviation variables 
meanandstdlocations <- intersect(grep("mean|std", measurementnames), grep("meanFreq", measurementnames, invert = TRUE))
#The above intentionally excludes the meanFreq measurements, as those are "Weighted average of the frequency components to obtain a mean frequency", which is distinct from the mean, which is Mean value, and is a member of the set for the same measurements such as Body Acceleration Jerk and Body Gyro.

# Create a reduced data out the subject and activity columns in preparation to reduce to only the mean and standard deviation measurements 
meanandstddata <- alldata[,-c(1:3)]

# Create a reduced data set containing only the mean and standard deviation measurements 
meanandstddata <- meanandstddata[,meanandstdlocations]

# Column bind the subjects and activities to the mean and stadard deviation measurements 
meanandstddata <- cbind(alldata[,c(1,3)], meanandstddata)

# Rename the first column in the mean and stadard deviation data set to a descriptive name 
colnames(meanandstddata)[1] <- "subject"

# Store the mean and stadard deviation data set column names to a vector 
thenames <- colnames(meanandstddata)

# Remove the leading digits and spaces from the names vector 
thenames <- gsub("\\d+ ","", thenames)

# Remove the opening parentheses from the names vector 
thenames <- gsub("\\(+","", thenames)

# Remove the closing parentheses from the names vector 
thenames <- gsub("\\)+","", thenames)

# Remove the dashes from in the from the names vector 
thenames <- gsub("-","", thenames)

# Substitute "acceleration" for "Acc: in the names vector 
thenames <- gsub("Acc","acceleration", thenames)

# Change the names vector to all lowercase (a preference listed durign the lecture by Dr. Leek) 
tolower(thenames)

# Rename the columns in the mean and stadard deviation data set 
colnames(meanandstddata) <- thenames

# create a tidy data set containing the average of each variable for each activity and each subject 
averagesdata <- meanandstddata %>% group_by(activityname, subject) %>% summarise_all("mean")

# Extract the variable names in the averages data set 
averagesnames <- colnames(averagesdata)

# Append "average" to each of the averages data names items 
averagesnames <- paste0(averagesnames,"average")

# Remove the "average" from activity name and subject, as these are not averages 
averagesnames[1:2] <- c("activityname", "subject")

# Update the variable names in the averages data set 
colnames(averagesdata) <- averagesnames

# check if the final step output file already exists, if so remove it
finalfile <- as.character(paste(startingwd, "measurementaveragesbysubjectandactivity.csv", sep = "/"))
invisible(if(file.exists(finalfile)) {file.remove(finalfile)})

# Write the final tidy data set to a csv and also View
write.csv(averagesdata, file = "measurementaveragesbysubjectandactivity.csv")
View(averagesdata)
```
